---
title: 'Pol Sci 200D Notes: Refresher'
author: 'Haotian (Barney) Chen'
date: "January 10, 2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

\newcommand{\y}{\mathbf{y}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\Cov}{\mathbb{Cov}}
\newcommand{\independent}{\perp\!\!\!\perp}

## OLS Refresher
In the classical regression model, we have the following assumptions:

+ (A.1) **Linearity**: $\Y = \X \mathbb{\beta} + \epsilon$
+ (A.2) **Strict Exogeneity**: $\E[\epsilon | \X] = 0$. The explanatory variables are exogenous (uncorrelated with the error term). It implies: 
  - (A.2.a) Weak Exogeneity: $\E[\X_i\epsilon_i] = 0$. The explanatory variable is orthogonal to the error term. 
  - (A.2.b) Unconditional Mean: $\E[\epsilon_i] = 0$. The error term has zero mean.
+ (A.3) **Full Rank**: $\text{rank}(\X)=k$ (we need $\X'\X$ to be invertible). A violation is *perfect multicollinearity* when some explanatory variables are linearly dependent.
+ (A.4) Spherical Errors: $\V[\epsilon | \X] = \E[\epsilon\epsilon'|\X] = \sigma^2 I_n$. It implies:
  - (A.4.a) **Homoskedasticity**: $\V[\epsilon_i | \X] = \sigma^2$. Error terms have constant variance. No heteroskedasticity. 
  - (A.4.b) **No Autocorrelation of Errors**: $\Cov[\epsilon_i, \epsilon_j | \X] = \E[\epsilon_i \epsilon_j | \X] = 0$ for $i \neq j$. No serial correlation.
+ (A.5) **Normality**: $\epsilon | \X \sim N(0, \sigma^2I_n)$. The error terms are normally distributed.

Under these assumptions, we can estimate the OLS estimator $\hat{\beta} = (\X'\X)^{-1}\X'Y$. Here are the properties of OLS Estimator: 

+ **Unbiasedness**: $\E[\hat{\beta} | \X] = \beta$. 
+ **Efficiency**: The Gauss–Markov theorem states that the OLS estimator is BLUE (Best Linear Unbiased Estimator). 
+ **Consistency**: $\hat{\beta} \xrightarrow{p} \beta$ as $n \rightarrow \infty$.
+ **Asymptotically Normality**: $\hat{\beta} \sim N(\beta, \sigma^2 (\X'\X)^{-1})$.

## Mathematics Refresher
This class assumes familiarity in probability theory and linear algebra. Check [Harvard's Math Prefresher](https://iqss.github.io/prefresher/) if you need a refresher. Below I provide a quick review.

+ **Prior Distribution**: the probability of the parameter $\theta$ before observing data, denoted as $\xi(\theta)$.
+ **Posterior Distribution**: the conditional distribution of $\theta$ given observed data $X_1, X_2, \ldots, X_n$, denoted as $\xi(\theta | x_1, \ldots, x_n)$.
+ **Likelihood Function**: $\xi(\theta | x_1, \ldots, x_n) \propto f_n(x_1, \ldots, x_n | \theta) \xi(\theta)$. Here the likelihood function is $L(\theta) = f_n(x_1, \ldots, x_n | \theta)$. 
+ Again: $\text{Posterior} \propto \text{Likelihood} \cdot \text{Prior}$
+ **Conjugate Families**: choosing a conjugate prior that the posterior is in same family as the prior. We discuss three conjugate families: Normal-Normal, Beta-Binomial, and Gamma-Poisson.
  - Normal Prior, Normal Data: 
```{r message=FALSE, warning=FALSE, echo=FALSE}
library(bayesrules)
plot_normal_normal(mean = 3.8, sd = 1.12, sigma = 5.8, y_bar = 3.35, n = 8)
```
  - Beta Prior, Binomial Data: 
```{r echo=FALSE}
plot_beta_binomial(alpha = 3, beta = 13, y = 5, n = 10) 
```
  - Gamma Prior, Poisson Data: 
```{r echo=FALSE}
plot_gamma_poisson(shape = 10, rate = 2, sum_y = 11, n = 4)
```

## Programming Refresher
We will use R programming language in this class and follow the [tidyverse coding style](https://style.tidyverse.org). We recommend RStudio but of course, you can pick your favorite IDE. You are expected to submit your problem sets and exams using Quarto -- check this [link](https://quarto.org/docs/faq/rmarkdown.html) for switching from RMarkdown to Quarto. For any output you produce for this class, make sure the tables (e.g., using `gt()` or `kable()`) and figures (using `ggplot`) are well-formatted and easy to read. 


### Simulation
Here is a quick review for Monte Carlo Simulation in R. 

```{r message=FALSE, warning=FALSE, eval=FALSE}
# common distributions (prefix r-; for density, prefix d-; for cumulative, prefix p-)
rnorm() # Normal
runif() # Uniform
rbinom() # Binomial
rpois() # Poisson
rexp() # Exponential

# apply family functions
apply() 
sapply()
lapply()
tapply()
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)

# random seeds for reproducibility
set.seed(2025)

# sample() function
sample(x = 1:10, size = 6, replace = FALSE)
sample(x = 1:10, size = 6, replace = TRUE)

# replicate() function
replicate(3, rnorm(2))

# map family functions (from purrr)
modtwo <- function(.x) {
  return(.x %% 2)
}

map(c(1, 8, 13, 20), modtwo) # apply modtwo() to each element in the vector

map_df(c(1, 8, 13, 20), function(.x) {
  return(data.frame(number = .x, 
                    mod2 = modtwo(.x)))
}) # apply modtwo() to each element in the vector and return a data frame

# simulate a simple dataset using function()
simulate_lm <- function(n = 100, beta0 = 2, beta1 = 3) {
  x <- rnorm(n)
  y <- beta0 + beta1 * x + rnorm(n)
  return(data.frame(x = x, y = y))
}

data <- simulate_lm(100, 2, 3)

ggplot(data, aes(x = x, y = y)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

